---
title: "Community Detection in R using communities of Friends characters"
author: "Keith McNulty"
date: "15/06/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this article I will use the community detection capabilities in the `igraph` package in R to show how to detect communities in a network.  By the end of the article we will able to see how the Louvain community detection algorithm breaks up the *Friends* characters into distinct communities (ignoring the obvious community of the six main characters), and if you are a fan of the show you can decide if this analysis makes sense to you.

## Data for assembling the network of Friends characters

In my earlier articles I showed how you can [scrape scripts of Friends episodes](https://drkeithmcnulty.com/2020/03/25/scraping-structured-data-from-semi-structured-documents/) and [use iterative programming](https://drkeithmcnulty.com/2020/03/30/simple-iterative-programming-and-error-handling-in-r/) to generate a network edgelist for the entire series of Friends.  A network edgelist is a simple dataset that contains the following:

1.  `from` and `to` columns to determine connections between character pairs in our (undirected) network, each character will be a node and each connection will be an edge.
2.  A `weight` column (which is a property of the edges) indicating the strength of the connection between the pair.  In this case this is determined by the number of different scenes the pair have appeared in together.

The edgelist I am using was generated using the techniques in the previous two articles, and loaded to Github in the [repo](https://github.com/keithmcnulty/friends_analysis) for this project.  I will now pull the dataset for all ten seasons down and we can take a look at it.

```{r libraries, message = FALSE, warning = FALSE}
library(tidyverse)
library(readr)
library(igraph)
```

```{r fetch data}

# get friends full series edgelist

edgefile_url <- "https://github.com/keithmcnulty/friends_analysis/blob/master/data/friends_full_series_edgelist.RDS?raw=true"
download.file(edgefile_url, "edgelist.RDS")

edgelist <- readRDS("edgelist.RDS")

knitr::kable(edgelist %>% head(10))

```

This looks as we expect.  So we are ready to start some work.

## Using the Louvain algorithm in `igraph` to find communities

Now, first, we are going to pretend that the six main characters don't know each other and delete all edges between them in our network.  This is because we are interested in how the other characters form communities around the main characters.  If we were to leave the main character connections intact, we know that they would form a very strong community amongst themselves, which was of course the whole point of the show.

```{r remove_six_friend_connections}
friends <- c("Phoebe", "Monica", "Rachel", "Joey", "Ross", "Chandler")
edgelist_without <- edgelist %>% 
  dplyr::filter(!(from %in% friends & to %in% friends))
```

Now we will convert our new edgelist into a matrix and then use that to build a graph object that has the `weight` column as a property of the edges:

``` {r}
edgelist_matrix <- as.matrix(edgelist_without[ ,c("from", "to")])
  
friends_graph <- igraph::graph_from_edgelist(edgelist_matrix, directed = FALSE) %>% 
  igraph::set.edge.attribute("weight", value = edgelist_without$weight)

```

We can now take a quick basic look at our *Friends* graph:

```{r pressure, echo=FALSE}
plot(friends_graph)
```

OK, what a mess - not surprising given that there are `r gorder(friends_graph)` vertices (characters) and `r gsize(friends_graph)` edges (connections) in this network.  We'll have to do some formatting for some nice plots at a later point.  But now we are ready to ask the Louvain algorithm to break this network into distinct communities.  The algorithm will try to maximize the strengths of connections inside a community and minimize connections between different communities. 

```{r run_louvain}
# run louvain with edge weights
louvain_partition <- igraph::cluster_louvain(friends_graph, weights = E(friends_graph)$weight)

# assign communities to graph
friends_graph$community <- louvain_partition$membership

# see how many communities there are

unique(friends_graph$community)

```

In looks like the algorithm found `r length(unique(friends_graph$community))` communities.  But we have no idea who is in them.  We can do a few things to try to understand each community better.  

1.  We can look at how big each community is.  Sometimes communities can be tiny and represent almost completely disconnected parts of a network (like a random scene between some characters that never appeared again).
2.  We can look at the "most important" person (vertex) in each community.  One way of doing this is to look for the vertex with the highest betweenness centrality, that is the person who connects the most characters in that community.

``` {r explore communities}
communities <- data.frame()

for (i in unique(friends_graph$community)) {
  # create subgraphs for each community
  subgraph <- induced_subgraph(friends_graph, v = which(friends_graph$community == i))
  # get size of each subgraph
  size <- igraph::gorder(subgraph)
  # get betweenness centrality
  btwn <-  igraph::betweenness(subgraph)
  communities <- communities %>% 
    dplyr::bind_rows(
      data.frame(community = i,
                 n_characters = size,
                 most_important = names(which(btwn == max(btwn)))
    )
  )
}

knitr::kable(communities %>% 
               dplyr::select(community, n_characters, most_important))

```

OK - we see there are a couple of communities that seem to be fairly small and probably quite disconnected (we will look at those in the appendix), but the main six communities orient around the six friends which is what we would expect.  This confirms that each of the six characters, despite their closeness to each other, also fostered fairly independent communities throughout the series.

We can try to get a sense of those communities by looking at the top five most important characters in each community (excluding the small communities).  This time we will look at a more simple measure of performance - the number of connections each character has, or their *degree* in the network.  

``` {r top_ten}

top_five <- data.frame()

for (i in unique(friends_graph$community)) {
# create subgraphs for each community
  subgraph <- induced_subgraph(friends_graph, v = which(friends_graph$community == i))
  
  # for larger communities
  if (igraph::gorder(subgraph) > 20) {
    # get degree
    degree <-  igraph::degree(subgraph)
    # get top ten degrees
    top <- names(head(sort(degree, decreasing = TRUE), 5))
    result <- data.frame(community = i, rank = 1:5, character = top)
  } else {
    result <- data.frame(community = NULL, rank = NULL, character = NULL)
  }
  
  top_five <- top_five %>% 
    dplyr::bind_rows(result)
}

knitr::kable(
  top_five %>% 
    tidyr::pivot_wider(names_from = rank, values_from = character)
)
```

We see some generic character names here like 'guy' or 'woman'.  If we ignore these, we can see the following communities of characters:

*  **Phoebe and the men in her life**:  Her eventual husband Mike, her drug-addicted massage client Steve, and her half-brother Frank (also her father's name)

*  **Monica, her parents and boyfriends**:  Chandler is missing here of course, by construction!

*  **Chandler and Janice**:  Outside of the six friends, Chandler has limited other connections with recurring characters outside of Janice.

*  **Joey and his acting connections**:  Outside of Gunther - who falls into Joey's community - Joey is mostly involved with directors and agents.

* **Rachel, her baby and her sister**:  Rachel's community orients mostly around the birth of her baby in Season 8.

* **Ross, Carol and Susan**:  Ross's community is dominated by the scenes with his ex-wife Carol and her girlfriend Susan, as well as his Paleontology professor girlfriend Charlie Wheeler.


## Visualizing the communities

We can now try to visualize these communities as part of the entire network.  To make it a bit easier to digest we are going to remove the labels for everyone except the six friends, and then we will color code the vertices and edges by community.  

```{r prep plot}
# give our nodes some properties, incl scaling them by degree and coloring them by community
  
V(friends_graph)$size <- 3
V(friends_graph)$frame.color <- "white"
V(friends_graph)$color <- friends_graph$community
V(friends_graph)$label <- V(friends_graph)$name
V(friends_graph)$label.cex <- 1.5

# also color edges according to their starting node
edge.start <- ends(friends_graph, es = E(friends_graph), names = F)[,1]
E(friends_graph)$color <- V(friends_graph)$color[edge.start]
E(friends_graph)$arrow.mode <- 0

# only label central characters

v_labels <- which(V(friends_graph)$name %in% friends)

for (i in 1:length(V(friends_graph))) {
  if (!(i %in% v_labels)) {
    V(friends_graph)$label[i] <- ""
  }
}


```

Now we can plot our graph.  The 'prettiest' plot is probably a spherical layout:

```{r plot sphere, fig.width = 10, fig.height = 10, fig.align = "center"}
l1 <- layout_on_sphere(friends_graph)
plot(friends_graph, rescale = T, layout = l1, main = "'Friends' Network - All Seasons")

```

But to better separate the communities for visual purposes, a force-directed plot is better:

```{r plot mds, fig.width = 10, fig.height = 10, fig.align = "center"}
l2 <- layout_with_mds(friends_graph)
plot(friends_graph, rescale = T, layout = l2, main = "'Friends' Network - All Seasons")

```


The second image is more helpful as it implies that Joey and Ross may lead more 'independent' lives and have more of a community of their own compared to the four other characters given that the force-directed algorithm has distanced them further from the others.

## Appendix:  Who are the smaller communities?

Let's have a look at the two smaller communities which popped up in our earlier analysis.  Let's look at who is in these communities.

```{r small communities}
small_communities <- data.frame()

for (i in unique(friends_graph$community)) {
# create subgraphs for each community
  subgraph <- induced_subgraph(friends_graph, v = which(friends_graph$community == i))
  
  # for larger communities
  if (igraph::gorder(subgraph) < 20) {
    # get degree
    degree <-  igraph::degree(subgraph)
    # get top ten degrees
    top <- names(sort(degree, decreasing = TRUE))
    result <- data.frame(community = i, rank = 1:length(top), character = top)
  } else {
    result <- data.frame(community = NULL, rank = NULL, character = NULL)
  }
  
  small_communities <- small_communities %>% 
    dplyr::bind_rows(result)
}

knitr::kable(
  small_communities %>% 
    tidyr::pivot_wider(names_from = rank, values_from = character)
)
```

Interestingly our algorithm appears to have picked up on a couple of specific episodes and regarded scenes inside them as disconnected networks of their own.  

*  The first of these communities appears to be from *The One Where Dr Remore Dies*
*  The second appears to be from *The One After I Do*
